---
title: "Spatial_HD_FALC"
author: "HWC"
date: "2026-09-03"
output:
  html_document:
    keep_md: true
    toc: true
    self_contained: true
  pdf_document:
    toc: true
editor_options:
  chunk_output_type: inline
---

Workflow: (1) SCT normalisation per sample, (2) cell-type signatures,
(3) z-score signature maps, (4) UCell scoring, (5) UCell maps and
winner-takes-all assignment, (6) zoomed views on the H&E.

```{r setup, include=FALSE}
options(width = 100)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, dev = c("png", "pdf"))
set.seed(1234)   # [CORRECTED] the seed was defined but never set
```

```{r library, include=FALSE}

library(Seurat)
library(SingleCellExperiment)
library(Matrix)
library(ggplot2)
library(patchwork)
library(UCell)
library(BiocParallel)
library(magick)
```


## Data input
```{r Data input}
data_dir <- "/R/Spatial_HD_2026/data/processed_seurat/"

seu_list <- list(
  obj1 = readRDS(file.path(data_dir, "430721_1-20260821_Hu_PericardFALC_AHP2_squA_S2_16um.rds")),
  obj2 = readRDS(file.path(data_dir, "433121_1-20260915_Hu_PericardFALC_AHP5_squA_S1_16um.rds")),
  obj3 = readRDS(file.path(data_dir, "433121_2-20260915_Hu_PericardFALC_AHP6_squD_S2_16um.rds")),
  obj4 = readRDS(file.path(data_dir, "430721_2-20260821_Hu_PericardFALC_AHP3_squD_S1_16um.rds"))
)
samples <- names(seu_list)

# full-resolution H&E images (same pixel space as the Seurat coordinates);
# samples without an entry fall back to the lowres image stored in the object
he_hires_path <- list(
  obj2 = "/R/Spatial_HD_2026/data/image to merge/20260916_Hu_pericard_AHP5PB1_STHD_squA.jpg"
)


for (s in samples) {
  message(s, ": ", ncol(seu_list[[s]]), " bins, ", nrow(seu_list[[s]]),
          " features; images: ", paste(Images(seu_list[[s]]), collapse = ", "))
}
```


## SCT normalization
```{r SCT normalization}
# After SCTransform the default assay is SCT, so in each SCE
#   counts    = SCT-corrected counts (used by UCell)
#   logcounts = log1p of those counts (used for gene maps and z-scores)
sce_list <- list()

for (s in samples) {
  seu   <- seu_list[[s]]
  assay <- DefaultAssay(seu)

  # [CORRECTED] layer= instead of the deprecated slot= (Seurat v5)
  seu <- seu[, colSums(GetAssayData(seu, assay = assay, layer = "counts")) > 0]
  seu <- SCTransform(seu, assay = assay, verbose = FALSE)

  # [SIMPLIFIED] removed the SCT_residuals assay and HVG annotations: nothing
  # downstream uses them, and the dense genes x bins residual matrix costs
  # several GB of memory per sample
  sce_list[[s]] <- as.SingleCellExperiment(seu)
  message(s, ": ", ncol(sce_list[[s]]), " bins retained")
}
```


## Spatial coordinate helpers
```{r Spatial coordinate helpers}
# ---------------------------------------------------------------------------
# get_coords(): the ONLY source of spatial coordinates in this document.
# Returns two frames for each SCE bin (same order as the SCE):
#   x, y   : lowres image pixels, y negated -> used by all whole-sample maps
#   px, py : full-resolution pixels         -> used by the zoomed H&E views
# ---------------------------------------------------------------------------
get_coords <- function(seu, sce, image_name = NULL) {
  if (is.null(image_name)) image_name <- Images(seu)[1]
  img <- seu[[image_name]]
  co  <- GetTissueCoordinates(img)

  bc   <- if ("cell" %in% colnames(co)) as.character(co$cell) else rownames(co)
  xcol <- if ("imagecol" %in% colnames(co)) "imagecol" else "x"
  ycol <- if ("imagerow" %in% colnames(co)) "imagerow" else "y"

  idx <- match(colnames(sce), bc)
  if (anyNA(idx))
    stop(sum(is.na(idx)), " SCE barcodes have no entry in the tissue coordinates.")

  px <- as.numeric(co[[xcol]])[idx]
  py <- as.numeric(co[[ycol]])[idx]

  he         <- GetImage(img, mode = "raster")
  lowres     <- ScaleFactors(img)$lowres
  is_fullres <- max(px, na.rm = TRUE) > ncol(he) * 1.05 ||
                max(py, na.rm = TRUE) > nrow(he) * 1.05
  f <- if (is_fullres) lowres else 1

  # [SIMPLIFIED] full-res pixels are returned here, so the zoom chunks no longer
  # re-read coordinates, re-match barcodes or test swap/flip orientations
  data.frame(x  =  px * f,
             y  = -py * f,
             px = if (is_fullres) px else px / lowres,
             py = if (is_fullres) py else py / lowres)
}

# crop_he(): H&E cropped to the tissue, in the same frame as x, y above
crop_he <- function(seu, image_name = NULL, pad_frac = 0.02) {
  if (is.null(image_name)) image_name <- Images(seu)[1]
  img <- seu[[image_name]]
  he  <- GetImage(img, mode = "raster")
  ih  <- nrow(he); iw <- ncol(he)

  co   <- GetTissueCoordinates(img)
  xcol <- if ("imagecol" %in% colnames(co)) "imagecol" else "x"
  ycol <- if ("imagerow" %in% colnames(co)) "imagerow" else "y"
  cx   <- as.numeric(co[[xcol]])
  cy   <- as.numeric(co[[ycol]])

  if (max(cx, na.rm = TRUE) > iw * 1.05 || max(cy, na.rm = TRUE) > ih * 1.05) {
    f  <- ScaleFactors(img)$lowres
    cx <- cx * f; cy <- cy * f
  }

  padx <- pad_frac * diff(range(cx, na.rm = TRUE))
  pady <- pad_frac * diff(range(cy, na.rm = TRUE))
  c1 <- max(1,  floor(min(cx, na.rm = TRUE) - padx))
  c2 <- min(iw, ceiling(max(cx, na.rm = TRUE) + padx))
  r1 <- max(1,  floor(min(cy, na.rm = TRUE) - pady))
  r2 <- min(ih, ceiling(max(cy, na.rm = TRUE) + pady))

  list(raster = he[r1:r2, c1:c2, drop = FALSE],
       xmin = c1, xmax = c2, ymin = -r2, ymax = -r1)
}

he_panel <- function(seu, label) {
  cr <- crop_he(seu)
  ggplot() +
    annotation_raster(cr$raster, xmin = cr$xmin, xmax = cr$xmax,
                      ymin = cr$ymin, ymax = cr$ymax) +
    xlim(cr$xmin, cr$xmax) + ylim(cr$ymin, cr$ymax) +
    coord_fixed() + theme_void() +
    labs(subtitle = paste0(label, " - H&E")) +
    theme(plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 10))
}

# image info per sample: lowres raster, scale factor and bin size
img_info <- lapply(seu_list, function(seu) {
  img <- seu[[Images(seu)[1]]]
  sf  <- ScaleFactors(img)
  list(raster = GetImage(img, mode = "raster"),
       lowres = sf$lowres,
       bin_px = sf$spot)          # bin width in full-res pixels
})

coord_list <- list()
he_panels  <- list()
for (s in samples) {
  coord_list[[s]] <- data.frame(sample = s, get_coords(seu_list[[s]], sce_list[[s]]))
  he_panels[[s]]  <- he_panel(seu_list[[s]], s)
}

wrap_plots(he_panels, nrow = 1)
```


## Plotting of selected genes
```{r Plotting of selected genes}
# [CORRECTED] unique(): CD14 was listed twice
genes_of_interest <- unique(c("PECAM1","CLDN5","ACKR1","VWF","PLVAP","ICAM2","AQP1","COL1A1","COL1A2","MYL9","LUM","ACTA2","PDGFRB","TAGLN","MYH11","CCL19","PDPN","VCAM1","IL33","CLU","IL7","CXCL12","DCN","LYZ","S100A9","S100A8","CTSS","CST3","GPNMB","CD68","MRC1","CD3D","CD3E","CD3G","IL7R","TRBC1","TRBC2","TRAC","MS4A1","CD79A","CD79B","IGHM","MZB1","XBP1","CD38","PROX1","LYVE1","VEGFR3"))

for (s in samples) {
  gone <- setdiff(genes_of_interest, rownames(sce_list[[s]]))
  if (length(gone)) message(s, " missing from SCT: ", paste(gone, collapse = ", "))
}

gene_plots <- list()

for (gene in genes_of_interest) {
  expr <- lapply(samples, function(s)
    if (gene %in% rownames(sce_list[[s]])) as.numeric(logcounts(sce_list[[s]])[gene, ]) else NULL)
  names(expr) <- samples
  det  <- vapply(expr, function(e) !is.null(e) && any(e > 0), logical(1))
  lims <- if (any(det)) range(unlist(expr[det])) else NULL

  panels <- lapply(samples, function(s) {
    d <- coord_list[[s]][, c("x", "y")]
    if (det[[s]]) {
      d$expression <- expr[[s]]
      d <- d[order(d$expression), ]
      ggplot(d, aes(x, y, color = expression)) +
        geom_point(size = 0.3, alpha = 0.6) +
        scale_color_viridis_c(option = "plasma", limits = lims) +
        coord_fixed() + theme_void() +
        labs(subtitle = s, color = "LogExpr") +
        theme(plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 10))
    } else {
      reason <- if (is.null(expr[[s]])) "not in dataset" else "all zero"
      ggplot(d, aes(x, y)) +
        geom_point(size = 0.3, alpha = 0.4, color = "grey80") +
        annotate("text", x = mean(range(d$x)), y = mean(range(d$y)),
                 label = paste0("Not detectable\n(", reason, ")"),
                 fontface = "italic", size = 4, color = "grey30") +
        coord_fixed() + theme_void() +
        labs(subtitle = s) +
        theme(plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 10))
    }
  })

  gene_plots[[gene]] <- wrap_plots(c(panels, he_panels), ncol = length(samples)) +
    plot_layout(guides = "collect") +
    plot_annotation(title = gene,
                    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14)))
  print(gene_plots[[gene]])
}
```


```{r Plotting of selected genes}
# [CORRECTED] unique(): CD14 was listed twice
genes_of_interest <- unique(c("PECAM1","PDGFRB","CCL19","PDPN","CXCL13","CCL21","CLU","CXCL12","DCN","CD3D","CD3E","CD3G","IL7R","LTA","LTB","LTBR","CCL21","TNFRSF11A","TNFSF11A","PROX1","LYVE1","VEGFR3"))

for (s in samples) {
  gone <- setdiff(genes_of_interest, rownames(sce_list[[s]]))
  if (length(gone)) message(s, " missing from SCT: ", paste(gone, collapse = ", "))
}

gene_plots <- list()

for (gene in genes_of_interest) {
  expr <- lapply(samples, function(s)
    if (gene %in% rownames(sce_list[[s]])) as.numeric(logcounts(sce_list[[s]])[gene, ]) else NULL)
  names(expr) <- samples
  det  <- vapply(expr, function(e) !is.null(e) && any(e > 0), logical(1))
  lims <- if (any(det)) range(unlist(expr[det])) else NULL

  panels <- lapply(samples, function(s) {
    d <- coord_list[[s]][, c("x", "y")]
    if (det[[s]]) {
      d$expression <- expr[[s]]
      d <- d[order(d$expression), ]
      ggplot(d, aes(x, y, color = expression)) +
        geom_point(size = 0.3, alpha = 0.6) +
        scale_color_viridis_c(option = "plasma", limits = lims) +
        coord_fixed() + theme_void() +
        labs(subtitle = s, color = "LogExpr") +
        theme(plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 10))
    } else {
      reason <- if (is.null(expr[[s]])) "not in dataset" else "all zero"
      ggplot(d, aes(x, y)) +
        geom_point(size = 0.3, alpha = 0.4, color = "grey80") +
        annotate("text", x = mean(range(d$x)), y = mean(range(d$y)),
                 label = paste0("Not detectable\n(", reason, ")"),
                 fontface = "italic", size = 4, color = "grey30") +
        coord_fixed() + theme_void() +
        labs(subtitle = s) +
        theme(plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 10))
    }
  })

  gene_plots[[gene]] <- wrap_plots(c(panels, he_panels), ncol = length(samples)) +
    plot_layout(guides = "collect") +
    plot_annotation(title = gene,
                    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14)))
  print(gene_plots[[gene]])
}
```

```{r Plotting of selected genes}
# [CORRECTED] unique(): CD14 was listed twice
genes_of_interest <- unique(c("RORC", "CCR6","CD7","KLRB1","ITGB7","SELE","SELPLG","IL6"))

for (s in samples) {
  gone <- setdiff(genes_of_interest, rownames(sce_list[[s]]))
  if (length(gone)) message(s, " missing from SCT: ", paste(gone, collapse = ", "))
}

gene_plots <- list()

for (gene in genes_of_interest) {
  expr <- lapply(samples, function(s)
    if (gene %in% rownames(sce_list[[s]])) as.numeric(logcounts(sce_list[[s]])[gene, ]) else NULL)
  names(expr) <- samples
  det  <- vapply(expr, function(e) !is.null(e) && any(e > 0), logical(1))
  lims <- if (any(det)) range(unlist(expr[det])) else NULL

  panels <- lapply(samples, function(s) {
    d <- coord_list[[s]][, c("x", "y")]
    if (det[[s]]) {
      d$expression <- expr[[s]]
      d <- d[order(d$expression), ]
      ggplot(d, aes(x, y, color = expression)) +
        geom_point(size = 0.3, alpha = 0.6) +
        scale_color_viridis_c(option = "plasma", limits = lims) +
        coord_fixed() + theme_void() +
        labs(subtitle = s, color = "LogExpr") +
        theme(plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 10))
    } else {
      reason <- if (is.null(expr[[s]])) "not in dataset" else "all zero"
      ggplot(d, aes(x, y)) +
        geom_point(size = 0.3, alpha = 0.4, color = "grey80") +
        annotate("text", x = mean(range(d$x)), y = mean(range(d$y)),
                 label = paste0("Not detectable\n(", reason, ")"),
                 fontface = "italic", size = 4, color = "grey30") +
        coord_fixed() + theme_void() +
        labs(subtitle = s) +
        theme(plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 10))
    }
  })

  gene_plots[[gene]] <- wrap_plots(c(panels, he_panels), ncol = length(samples)) +
    plot_layout(guides = "collect") +
    plot_annotation(title = gene,
                    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14)))
  print(gene_plots[[gene]])
}
```




## Set up cell type signatures
```{r Set up cell types}
my_signatures <- list(
  VEC         = c("PECAM1","CLDN5","ACKR1","VWF","PLVAP","ICAM2","AQP1"),
  LEC         = c("PECAM1","PROX1","LYVE1","VEGFR3"),
  PeriFb_VSMC = c("COL1A1","COL1A2","MYL9","LUM","ACTA2","PDGFRB","TAGLN","MYH11"),
  FRC         = c("CCL19","PDPN","VCAM1","IL33","CLU","IL7","CXCL12","DCN"),
  Myeloid     = c("LYZ","S100A9","S100A8","CTSS","CST3","GPNMB","CD68","MRC1"),
  Tcell       = c("CD3D","CD3E","CD3G","IL7R","TRBC1","TRBC2","TRAC"),
  Bcell       = c("MS4A1","CD79A","CD79B","IGHM","MZB1","XBP1","CD38")
)

# genes must be present in EVERY object so scores use the same gene set
common_genes <- Reduce(intersect, lapply(sce_list, rownames))
sig_present  <- lapply(my_signatures, intersect, common_genes)

for (sig in names(my_signatures)) {
  missing <- setdiff(my_signatures[[sig]], sig_present[[sig]])
  message(sig, ": ", length(sig_present[[sig]]), "/", length(my_signatures[[sig]]), " present",
          if (length(missing)) paste0("  MISSING: ", paste(missing, collapse = ", ")) else "")
}

sig_use <- sig_present[lengths(sig_present) >= 3]
message("Using ", length(sig_use), " signatures: ", paste(names(sig_use), collapse = ", "))

cell_colors <- c(
  VEC         = "#ff7f0e",
  LEC         = "#511cd8",
  PeriFb_VSMC = "#2ca02c",
  FRC         = "#E41A1C",
  Myeloid     = "#e377c2",
  Tcell       = "#bcbd22",
  Bcell       = "#17becf",
  Unassigned  = "grey90"
)

# [SIMPLIFIED] one map function for every signature-score plot below
plot_score_maps <- function(scores, sig, colour_scale, legend_title) {
  panels <- lapply(samples, function(s) {
    d <- data.frame(x = coord_list[[s]]$x, y = coord_list[[s]]$y,
                    score = scores[[s]][, sig])
    d <- d[order(d$score), ]
    ggplot(d, aes(x, y, color = score)) +
      geom_point(size = 0.4, alpha = 0.8) +
      colour_scale +
      coord_fixed() + theme_void() +
      labs(subtitle = s, color = legend_title) +
      theme(plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 10))
  })
  wrap_plots(c(panels, he_panels), ncol = length(samples)) +
    plot_layout(guides = "collect") +
    plot_annotation(title = sig,
                    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14)))
}
```


## Z-score signature maps
```{r Z-score signature maps}
score_signature <- function(sce, genes) {
  mat <- as.matrix(logcounts(sce)[genes, , drop = FALSE])
  mu  <- rowMeans(mat)
  sdv <- apply(mat, 1, sd)
  sdv[sdv == 0 | is.na(sdv)] <- 1            # zero-variance genes -> score 0
  colMeans((mat - mu) / sdv)
}

zscore_list <- lapply(sce_list, function(sce)
  sapply(sig_use, function(g) score_signature(sce, g)))

keep_z <- Reduce(intersect, lapply(zscore_list, function(m) colnames(m)[apply(m, 2, sd) > 0]))
zscore_list <- lapply(zscore_list, function(m) m[, keep_z, drop = FALSE])

zscore_plots <- list()
for (sig in keep_z) {
  v <- unlist(lapply(zscore_list, function(m) m[, sig]))
  zscore_plots[[sig]] <- plot_score_maps(
    zscore_list, sig,
    scale_color_viridis_c(option = "magma", limits = quantile(v, c(0.01, 0.99)),
                          oob = scales::squish),
    "z-score")
  print(zscore_plots[[sig]])
}
```


## UCell scoring
```{r UCell scoring}
score_ucell <- function(sce, sigs, maxRank, ncores = 4) {
  cnt <- counts(sce)
  if (is.null(colnames(cnt)) || anyDuplicated(colnames(cnt)))
    stop("Bins need unique colnames to keep scores aligned with coordinates")

  sigs <- lapply(sigs, intersect, rownames(cnt))
  sigs <- sigs[lengths(sigs) > 0]

  sc <- ScoreSignatures_UCell(matrix = cnt, features = sigs, maxRank = maxRank,
                              name = "", chunk.size = 2000,
                              BPPARAM = MulticoreParam(ncores))   # SnowParam() on Windows
  sc[colnames(sce), names(sigs), drop = FALSE]                    # enforce bin order
}

# maxRank: a signature gene only counts if it ranks within the top max_rank
# genes of a bin. Set from the data (90th percentile of detected genes per bin).
detected <- unlist(lapply(sce_list, function(sce) colSums(counts(sce) > 0)))
print(summary(detected))
max_rank <- max(100, ceiling(quantile(detected, 0.9)))

score_list <- lapply(sce_list, score_ucell, sigs = sig_use, maxRank = max_rank)
for (s in samples) stopifnot(nrow(score_list[[s]]) == nrow(coord_list[[s]]))

keep <- Reduce(intersect, lapply(score_list, function(m) colnames(m)[apply(m, 2, sd) > 0]))
score_list <- lapply(score_list, function(m) m[, keep, drop = FALSE])
message("Retained: ", paste(keep, collapse = ", "))

cts <- intersect(keep, names(cell_colors))
if (length(setdiff(keep, cts)))
  message("No colour defined, skipping: ", paste(setdiff(keep, cts), collapse = ", "))
```


## UCell score maps
```{r UCell score maps}

ucell_plots <- list()
for (ct in cts) {
  v <- unlist(lapply(score_list, function(m) m[, ct]))
  ucell_plots[[ct]] <- plot_score_maps(
    score_list, ct,
    scale_color_gradient(low = "grey92", high = cell_colors[[ct]],
                         limits = c(0, quantile(v, 0.99)), oob = scales::squish),
    "UCell score")
  print(ucell_plots[[ct]])
}
```


## Winner-takes-all assignment
```{r Winner-takes-all}
min_score  <- 0       # Unassigned only if every signature scores 0
min_margin <- 0       # e.g. 0.02 = top score must beat the runner-up by this much
scale_sigs <- FALSE   # TRUE = rescale each signature by its pooled 99th percentile

sig_ref <- sapply(cts, function(ct)
  quantile(unlist(lapply(score_list, function(m) m[, ct])), 0.99))
sig_ref[sig_ref == 0] <- 1

assign_winner <- function(m) {
  m <- m[, cts, drop = FALSE]
  if (scale_sigs) m <- sweep(m, 2, sig_ref[cts], "/")

  idx    <- max.col(m, ties.method = "first")
  top    <- m[cbind(seq_len(nrow(m)), idx)]
  m2     <- m
  m2[cbind(seq_len(nrow(m)), idx)] <- -Inf
  second <- do.call(pmax, as.data.frame(m2))

  label <- colnames(m)[idx]
  label[top <= min_score | (top - second) < min_margin] <- "Unassigned"

  data.frame(celltype  = factor(label, levels = c(cts, "Unassigned")),
             top_score = top,
             margin    = top - second)
}

winner_list <- lapply(score_list, assign_winner)

assign_df <- do.call(rbind, lapply(samples, function(s)
  data.frame(coord_list[[s]], label = winner_list[[s]]$celltype)))

print(sapply(winner_list, function(w) round(100 * prop.table(table(w$celltype)), 1)))

ct_cols <- c(cell_colors[cts], Unassigned = "grey90")

panels <- lapply(samples, function(s) {
  d <- assign_df[assign_df$sample == s, ]
  d <- d[order(d$label != "Unassigned"), ]          # Unassigned drawn first
  ggplot(d, aes(x, y, color = label)) +
    geom_point(size = 0.4, alpha = 0.9) +
    scale_color_manual(values = ct_cols, drop = FALSE, name = "Cell type") +
    guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
    coord_fixed() + theme_void() +
    labs(subtitle = s) +
    theme(plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 10))
})

p_winner <- wrap_plots(c(panels, he_panels), ncol = length(samples)) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Winner-takes-all cell type assignment (UCell)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
print(p_winner)
```


## Cell type distribution
```{r Cell type distribution}
# one cell type highlighted, all other bins grey
highlight_plots <- list()

for (ct in cts) {
  cols <- setNames(rep("grey85", nlevels(assign_df$label)), levels(assign_df$label))
  cols[ct] <- cell_colors[[ct]]

  panels <- lapply(samples, function(s) {
    d <- assign_df[assign_df$sample == s, ]
    d <- d[order(d$label == ct), ]                    # target drawn on top
    n <- sum(d$label == ct)
    ggplot(d, aes(x, y, color = label)) +
      geom_point(size = 0.15) +
      scale_color_manual(values = cols, drop = FALSE, guide = "none") +
      coord_fixed() + theme_void() +
      labs(subtitle = sprintf("%s\n%s bins (%s%%)", s, format(n, big.mark = ","),
                              round(100 * n / nrow(d), 1))) +
      theme(plot.subtitle = element_text(hjust = 0.5, face = "bold"))
  })

  highlight_plots[[ct]] <- wrap_plots(panels, nrow = 1) +
    plot_annotation(title = sprintf("Spatial distribution: %s (one cell type per bin)", ct),
                    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16)))
  print(highlight_plots[[ct]])
}
```

```{r Selected cell types together}

sel_types <- c("Bcell", "FRC", "LEC")
stopifnot(all(sel_types %in% cts))

cols <- setNames(rep("grey85", nlevels(assign_df$label)), levels(assign_df$label))
cols[sel_types] <- cell_colors[sel_types]

panels <- lapply(samples, function(s) {
  d <- assign_df[assign_df$sample == s, ]
  d <- d[order(d$label %in% sel_types), ]
  # [CORRECTED] counts for every selected type (only T cell and FRC were shown)
  counts_txt <- paste(sprintf("%s: %s", sel_types,
                              format(sapply(sel_types, function(ct) sum(d$label == ct)),
                                     big.mark = ",")), collapse = " | ")
  ggplot(d, aes(x, y, color = label)) +
    geom_point(size = 0.35) +
    scale_color_manual(values = cols, breaks = sel_types, name = "Cell type") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    coord_fixed() + theme_void() +
    labs(subtitle = paste0(s, "\n", counts_txt)) +
    theme(plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 8))
})

p_selected <- wrap_plots(panels, nrow = 1) +
  plot_layout(guides = "collect") +
  plot_annotation(title = paste(sel_types, collapse = ", "),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16)))
print(p_selected)
```


## Zoomed views on the H&E
```{r Zoom settings}
zoom_sample  <- "obj2"
zoom_windows <- list(                       # in coord_list units (x, y)
  zoom1 = list(x = c(100, 265), y = c(-405, -295)),
  zoom2 = list(x = c(145, 205), y = c(-400, -340))
)

# read each full-res H&E once (slow; ~20k x 18k px)
he_full <- lapply(he_hires_path, image_read)

panel_theme <- theme_void() +
  theme(plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 11),
        panel.border  = element_rect(color = "black", fill = NA, linewidth = 0.5))

# overview with the zoom windows, used to choose / check the coordinates
ggplot(coord_list[[zoom_sample]], aes(x, y)) +
  geom_point(size = 0.1, color = "grey60") +
  lapply(names(zoom_windows), function(w)
    annotate("rect", xmin = zoom_windows[[w]]$x[1], xmax = zoom_windows[[w]]$x[2],
             ymin = zoom_windows[[w]]$y[1], ymax = zoom_windows[[w]]$y[2],
             fill = NA, color = "black")) +
  coord_fixed() + theme_minimal() +
  labs(title = paste(zoom_sample, "- zoom windows"))
```

```{r Zoom helpers}
# H&E for a window given in full-res pixels: hi-res JPG if available,
# otherwise the lowres image stored in the Seurat object
he_layer_fullres <- function(s, xl, yl, crop_width = 2000) {
  if (!is.null(he_full[[s]])) {
    geom <- sprintf("%dx%d+%d+%d", ceiling(diff(xl)), ceiling(diff(yl)),
                    floor(xl[1]), floor(yl[1]))
    crop <- image_crop(he_full[[s]], geom)
    # [CORRECTED] downscaling is applied to the crop that is plotted; before,
    # image_scale() ran after the layer was built and had no effect
    if (image_info(crop)$width > crop_width)
      crop <- image_scale(crop, as.character(crop_width))
    ras <- as.raster(crop)
  } else {
    f  <- img_info[[s]]$lowres
    he <- img_info[[s]]$raster
    r1 <- max(1, floor(yl[1] * f));  r2 <- min(nrow(he), ceiling(yl[2] * f))
    c1 <- max(1, floor(xl[1] * f));  c2 <- min(ncol(he), ceiling(xl[2] * f))
    ras <- he[r1:r2, c1:c2]
    xl  <- c(c1 - 1, c2) / f
    yl  <- c(r1 - 1, r2) / f
  }
  annotation_raster(ras, xmin = xl[1], xmax = xl[2], ymin = -yl[2], ymax = -yl[1])
}

# bins, limits, H&E layer and coordinate system for one zoom window
zoom_frame <- function(s, win, crop_width = 2000) {
  d <- data.frame(coord_list[[s]], label = winner_list[[s]]$celltype)
  in_win <- d$x >= win$x[1] & d$x <= win$x[2] & d$y >= win$y[1] & d$y <= win$y[2]
  if (!any(in_win)) stop("No bins inside the zoom window")

  b  <- img_info[[s]]$bin_px
  xl <- range(d$px[in_win]) + c(-b, b) / 2         # include the full edge bins
  yl <- range(d$py[in_win]) + c(-b, b) / 2

  list(bins = d[in_win, ], in_win = in_win, b = b,
       he = he_layer_fullres(s, xl, yl, crop_width),
       coord = coord_fixed(xlim = xl, ylim = -rev(yl), expand = FALSE))
}

# cell types on the H&E: one panel per type (others grey) + merged panel
plot_zoom <- function(s, win, types = cts, alpha_target = 1, alpha_other = 0.6,
                      other_col = "grey80", crop_width = 2000, ncol = 4, title = "zoom") {
  z  <- zoom_frame(s, win, crop_width)
  dz <- z$bins

  panel <- function(sel = NULL, subtitle, legend = FALSE) {
    p <- ggplot(dz, aes(px, -py)) + geom_blank() + z$he
    if (length(sel)) {
      tgt <- dz$label %in% sel
      # [CORRECTED] tiles of the exact bin size: points sized in mm overlapped
      # and hid neighbouring bins in the merged panel
      p <- p +
        geom_tile(data = dz[!tgt, ], fill = other_col,
                  width = z$b, height = z$b, alpha = alpha_other) +
        geom_tile(data = dz[tgt, ], aes(fill = label),
                  width = z$b, height = z$b, alpha = alpha_target) +
        scale_fill_manual(values = cell_colors[types], limits = types, name = "Cell type",
                          guide = if (legend) "legend" else "none")
    }
    p + z$coord + panel_theme + labs(subtitle = subtitle)
  }

  panels <- c(
    list(panel(subtitle = "H&E")),
    lapply(types, function(ct)
      panel(ct, sprintf("%s\n(n = %s)", ct, format(sum(dz$label == ct), big.mark = ",")))),
    list(panel(types, "All cell types", legend = TRUE))
  )

  wrap_plots(panels, ncol = ncol) +
    plot_layout(guides = "collect") +
    plot_annotation(title = sprintf("%s: cell types on H&E (%s)", s, title),
                    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10)))
}

# marker genes on the H&E: whole captured area (top row) and zoom (bottom row)
plot_marker_zoom <- function(s, win, markers, marker_cols, alpha_range = c(0.3, 1),
                             crop_width = 2000) {
  sce     <- sce_list[[s]]
  markers <- intersect(markers, rownames(sce))
  lc      <- as.matrix(logcounts(sce)[markers, , drop = FALSE])
  markers <- markers[rowSums(lc > 0) > 0]
  if (!length(markers)) stop("None of the markers is detected in ", s)

  d <- data.frame(coord_list[[s]][, c("x", "y", "px", "py")],
                  t(lc[markers, , drop = FALSE]), check.names = FALSE)
  d$ny <- -d$py
  q99  <- sapply(markers, function(g) unname(quantile(d[[g]][d[[g]] > 0], 0.99)))

  z  <- zoom_frame(s, win, crop_width)
  cr <- crop_he(seu_list[[s]])

  views <- list(
    Full = list(rows = rep(TRUE, nrow(d)), xc = "x", yc = "y",
                w = img_info[[s]]$bin_px * img_info[[s]]$lowres,
                he = annotation_raster(cr$raster, xmin = cr$xmin, xmax = cr$xmax,
                                       ymin = cr$ymin, ymax = cr$ymax),
                coord = coord_fixed(xlim = c(cr$xmin, cr$xmax), ylim = c(cr$ymin, cr$ymax),
                                    expand = FALSE),
                box = TRUE),
    Zoom = list(rows = z$in_win, xc = "px", yc = "ny", w = z$b,
                he = z$he, coord = z$coord, box = FALSE)
  )

  panel <- function(v, genes, subtitle) {
    view  <- views[[v]]
    frame <- d[view$rows, ]
    p <- ggplot(frame, aes(.data[[view$xc]], .data[[view$yc]])) + geom_blank() + view$he
    for (g in genes) {
      dd <- frame[frame[[g]] > 0, ]
      dd$gene <- g
      dd$a    <- pmin(dd[[g]] / q99[[g]], 1)
      p <- p + geom_tile(data = dd,
                         aes(.data[[view$xc]], .data[[view$yc]], fill = gene, alpha = a),
                         width = view$w, height = view$w, inherit.aes = FALSE)
    }
    if (view$box)
      p <- p + annotate("rect", xmin = win$x[1], xmax = win$x[2],
                        ymin = win$y[1], ymax = win$y[2],
                        fill = NA, color = "black", linewidth = 0.6)
    p +
      scale_fill_manual(values = marker_cols[markers], limits = markers, name = "Marker",
                        guide = if (length(genes) == length(markers))
                          guide_legend(override.aes = list(alpha = 1)) else "none") +
      scale_alpha_continuous(range = alpha_range, limits = c(0, 1), guide = "none") +
      view$coord + panel_theme + labs(subtitle = subtitle)
  }

  panels <- list()
  for (v in names(views)) {
    panels[[paste(v, "HE")]] <- panel(v, character(0), paste(v, "- H&E"))
    for (g in markers) panels[[paste(v, g)]] <- panel(v, g, paste(v, "-", g))
    panels[[paste(v, "merged")]] <- panel(v, markers, paste(v, "- merged"))
  }

  wrap_plots(panels, ncol = length(markers) + 2) +
    plot_layout(guides = "collect") +
    plot_annotation(title = sprintf("%s: %s on H&E", s, paste(markers, collapse = ", ")),
                    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10)))
}
```

```{r Zoom 1, fig.width = 16, fig.height = 8}
p_zoom1 <- plot_zoom(zoom_sample, zoom_windows$zoom1, title = "zoom 1")
print(p_zoom1)
```

```{r Zoom 2, fig.width = 16, fig.height = 8}
# smaller window for nuclear detail: more transparent tiles, sharper crop
p_zoom2 <- plot_zoom(zoom_sample, zoom_windows$zoom2,
                     alpha_target = 0.5, alpha_other = 0.3, crop_width = 2500,
                     title = "zoom 2")
print(p_zoom2)
```


```{r Marker genes on H&E, fig.width = 20, fig.height = 8}
marker_window1 <- zoom_windows$zoom1      # or zoom_windows$zoom2 for the smaller window

p_markers_acta2 <- plot_marker_zoom(
  zoom_sample, marker_window1,
  markers     = c("CCL19", "IL7R", "ACTA2"),
  marker_cols = c(CCL19 = "#FF1493", IL7R = "#00BFFF", ACTA2 = "#32CD32")
)
print(p_markers_acta2)



marker_window2 <- zoom_windows$zoom2      # or zoom_windows$zoom2 for the smaller window
p_markers_acta2 <- plot_marker_zoom(
  zoom_sample, marker_window2,
  markers     = c("CCL19", "IL7R", "ACTA2"),
  marker_cols = c(CCL19 = "#FF1493", IL7R = "#00BFFF", ACTA2 = "#32CD32")
)
print(p_markers_acta2)

```


## Session info
```{r date and session info}
date()
sessionInfo()
```
